Nonlinear Dendritic Coincidence Detection for Supervised Learning

### 2021-03-29 notes/draft_article.tex
.

		In a supervised learning scheme, where the top down input
		arriving at the apical compartment acts as the teaching signal,
		the most straight-forward learning rule for the basal synaptic
		weights would be derived from an appropriate loss function,
		such as a mean square error, based on the difference between 
		basal and apical input, i.e. $I_p - I_d$. Theoretical work has 
		investigated possible learning mechanisms
		that could utilize an intracellular error signal
		\citep{Urbanczik_2014,Schiess_2016,Guerguiev_2017}.
		However, a clear experimental
		evidence for a physical quantity encoding such an error 
		is---to our knowledge---yet to be found. 
		On the other hand, Hebbian-type plasticity is extensively
		documented in experiments. Therefore, our work was
		based on the question whether the non-linear interactions
		between basal and apical synaptic input could, when
		combined with a Hebbian plasticity rule, allow a neuron
		to learn to reproduce an apical teaching signal in its
		proximal input.

@@

		The neuron model used throughout this study is a discrete-time rate encoding model that
		uses two separate input variables, subsuming the total synaptic
		input current injected arriving at the basal (proximal) and apical (distal) 
		dendritic structure of a pyramidal neuron, respectively. The model is a slightly
		simplified version of a phenomenological model proposed by \citet{Shai_2015}.
		Denoting the input currents by $I_p$ (proximal) and $I_d$ (distal), 
		the model is written as
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{align}
		\begin{split}
		y\left(t\right) &= \sigma\left( I_p(t) - \theta_{p0} \right)
		\left[1-\sigma\left(I_d(t) - \theta_d\right)\right] \\
		&+ \alpha \sigma\left(I_d(t) - \theta_d \right)\sigma\left( I_p(t) - \theta_{p1} \right)
		\end{split} \label{eq:comp_model}\\
		\sigma(x) &\equiv \frac{1}{1+\exp(-4x)} \; .
		\end{align}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		Here, $\theta_{p0}$, $\theta_{p1}$ and $\theta_d$ are threshold variables with
		respect to proximal and distal input. Overall, this equation describes two distinct
		regions of neural activation in the $(I_p, I_d)$-space which differ in their
		maximal firing rates, which are set to $1$ and $\alpha$, where $0 < \alpha < 1$.
		A plot of \eqref{eq:comp_model} is shown in Fig.~\ref{fig:comp_model}.
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{figure}
			\centering
			\includegraphics[width=0.8\columnwidth]{plot_comp_mod_marks.png}
			\caption{{\bf Two-compartment rate model.} Firing rate as a
			function of proximal and distal input $I_p$ and $I_d$, 
			see \eqref{eq:comp_model}. 
			The thresholds $\theta_{p0}$, $\theta_{p1}$ and $\theta_d$ define
			two regions of neural activity with a a maximal firing rate of $1$
			and $\alpha$.}
			\label{fig:comp_model}
		\end{figure}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

		In all our numerical experiments, we compared this model with a 
		simple point neuron model that was given by
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{equation}
			y(t) = \sigma\left(I_p(t) + I_d(t) - \theta \right) \; .
		\end{equation}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		In all our numerical experiments, the apical input $I_d$ was generated
		``as is", meaning, it was not dynamically calculated as a superposition
		of multiple presynaptic inputs, but given by
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{equation}
			I_d(t) = n_d(t) x_d(t) - b_d(t) \; ,
			\label{eq:I_d}
		\end{equation}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		where $n_d(t)$ is a scaling factor, $x_d(t)$ a pre-generated
		discrete time sequence and $b_d(t)$ a bias. Note that $n_d$ and $b_d$ 
		are time dependent since they were subject to adaptation processes 
		that are described in the next section.

		Similarly, $I_p(t)$ was given by
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{equation}
			I_p(t) = n_p(t) \sum_{i=1}^{N} x_{p,i}(t) w_i(t) - b_p(t) \; ,
		\end{equation}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		where $N$ is the number of presynaptic afferents, $x_{p,i}(t)$ the
		corresponding sequences and $w_i(t)$ the synaptic efficacies.
		As for $I_d(t)$, $n_p(t)$ and $b_p(t)$ was used as a time dependent
		scaling and bias.

@@

		Additionally, the scaling and bias variables were changing dynamically
		according to the following homeostatic plasticity rules:
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{align}
			b_p(t+1) &= b_p(t) + \mu_b \left[I_p(t) - I_p^t\right] \\
			b_d(t+1) &= b_d(t) + \mu_b \left[I_d(t) - I_d^t\right] \\
			n_p(t+1) &= n_p(t) + \mu_n \left[V_p^t - \left( I_p(t) - \tilde{I}_p(t)\right)^2\right] \\
			n_d(t+1) &= n_d(t) + \mu_n \left[V_d^t - \left( I_d(t) - \tilde{I}_d(t)\right)^2\right] \\
			\tilde{I}_p(t+1) &= (1-\mu_{av})\tilde{I}_p(t) + \mu_{av} I_p(t+1) \\
			\tilde{I}_d(t+1) &= (1-\mu_{av})\tilde{I}_d(t) + \mu_{av} I_d(t+1)
		\end{align}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		Here, $I_p^t$, $I_d^t$, $V_p^t$ and $V_p^t$ define targets for the 
		temporal means and variances of $I_p$ and $I_d$. The dynamic variables 
		$\tilde{I}_p$ and $\tilde{I}_d$ are simply low-pass filtered running 
		averages of $I_p$ and $I_d$.

@@

		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{table}
			\begin{tabular}{ l | l || l | l }
				$\theta_{p0}$ & $0$ & $V_d^t$ & $0.25$ \\
				$\theta_{p1}$ & $-1$ & $\mu_b$ & $10^{-3}$ \\ 
				$\theta_{d}$ & $0$ & $\mu_n$ & $10^{-4}$ \\  
				$\alpha$ & $0.3$ & $\mu_{av}$ & $5 \cdot 10^{-3}$ \\   
				$\mu_w$ & $5 \cdot 10^{-5}$ & $I_p^t$ & $0$ \\
				$V_p^t$ & $0.25$ & $I_d^t$ & $0$  
			\end{tabular}
		\caption{Model parameters}
		\label{tab:parameters}
		\end{table}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

		\section{Results}

		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{figure}
			\includegraphics[width=\columnwidth]{pd_corr_scaling}
		\end{figure}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{figure}
			\includegraphics[width=\columnwidth]{distraction_classification_compare}
		\end{figure}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

		In a supervised learning scheme, where the top down input
		arriving at the apical compartment acts as the teaching signal,
		the most straight-forward learning rule for the basal synaptic
		weights would be derived from an appropriate loss function,
		such as a mean square error, based on the difference between 
		basal and apical input, i.e. $I_{\rm p} - I_{\rm d}$. Theoretical work has 
		investigated possible learning mechanisms
		that could utilize an intracellular error signal
		\citep{Urbanczik_2014,Schiess_2016,Guerguiev_2017}.
		However, a clear experimental
		evidence for a physical quantity encoding such an error 
		is---to our knowledge---yet to be found. 
		On the other hand, Hebbian-type plasticity is extensively
		documented in experiments. Therefore, our work was
		based on the question whether the non-linear interactions
		between basal and apical synaptic input could, when
		combined with a Hebbian plasticity rule, allow a neuron
		to learn to reproduce an apical teaching signal in its
		proximal input.

@@

		The neuron model used throughout this study is a discrete-time rate encoding model that
		uses two separate input variables, subsuming the total synaptic
		input current injected arriving at the basal (proximal) and apical (distal) 
		dendritic structure of a pyramidal neuron, respectively. The model is a slightly
		simplified version of a phenomenological model proposed by \citet{Shai_2015}.
		Denoting the input currents by $I_{\rm p}$ (proximal) and $I_{\rm d}$ (distal), 
		the model is written as
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{align}
		\begin{split}
		y\left(t\right) &= \sigma\left( I_{\rm p}(t) - \theta_{p0} \right)
		\left[1-\sigma\left(I_{\rm d}(t) - \theta_d\right)\right] \\
		&+ \alpha \sigma\left(I_{\rm d}(t) - \theta_d \right)\sigma\left( I_{\rm p}(t) - \theta_{p1} \right)
		\end{split} \label{eq:comp_model}\\
		\sigma(x) &\equiv \frac{1}{1+\exp(-4x)} \; .
		\end{align}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		Here, $\theta_{p0}$, $\theta_{p1}$ and $\theta_d$ are threshold variables with
		respect to proximal and distal input. Overall, this equation describes two distinct
		regions of neural activation in the $(I_{\rm p}, I_{\rm d})$-space which differ in their
		maximal firing rates, which are set to $1$ and $\alpha$, where $0 < \alpha < 1$.
		A plot of \eqref{eq:comp_model} is shown in Fig.~\ref{fig:comp_model}.
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{figure}
			\centering
			\includegraphics[width=0.8\columnwidth]{plot_comp_mod_marks.png}
			\caption{{\bf Two-compartment rate model.} Firing rate as a
			function of proximal and distal input $I_{\rm p}$ and $I_{\rm d}$, 
			see \eqref{eq:comp_model}. 
			The thresholds $\theta_{p0}$, $\theta_{p1}$ and $\theta_d$ define
			two regions of neural activity with a a maximal firing rate of $1$
			and $\alpha$.}
			\label{fig:comp_model}
		\end{figure}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

		In all our numerical experiments, we compared this model with a 
		simple point neuron model that was given by
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{equation}
			y(t) = \sigma\left(I_{\rm p}(t) + I_{\rm d}(t) - \theta \right) \; .
		\end{equation}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		In all our numerical experiments, the apical input $I_{\rm d}$ was generated
		``as is", meaning, it was not dynamically calculated as a superposition
		of multiple presynaptic inputs, but given by
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{equation}
			I_{\rm d}(t) = n_d(t) x_d(t) - b_d(t) \; ,
			\label{eq:I_d}
		\end{equation}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		where $n_d(t)$ is a scaling factor, $x_d(t)$ a pre-generated
		discrete time sequence and $b_d(t)$ a bias. Note that $n_d$ and $b_d$ 
		are time dependent since they were subject to adaptation processes 
		that are described in the next section.

		Similarly, $I_{\rm p}(t)$ was given by
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{equation}
			I_{\rm p}(t) = n_p(t) \sum_{i=1}^{N} x_{p,i}(t) w_i(t) - b_p(t) \; ,
		\end{equation}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		where $N$ is the number of presynaptic afferents, $x_{p,i}(t)$ the
		corresponding sequences and $w_i(t)$ the synaptic efficacies.
		As for $I_{\rm d}(t)$, $n_p(t)$ and $b_p(t)$ was used as a time dependent
		scaling and bias.

@@

		Additionally, the scaling and bias variables were changing dynamically
		according to the following homeostatic plasticity rules:
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{align}
			b_p(t+1) &= b_p(t) + \mu_b \left[I_{\rm p}(t) - I_{\rm p}^t\right] \\
			b_d(t+1) &= b_d(t) + \mu_b \left[I_{\rm d}(t) - I_{\rm d}^t\right] \\
			n_p(t+1) &= n_p(t) + \mu_n \left[V_p^t - \left( I_{\rm p}(t) - \tilde{I}_p(t)\right)^2\right] \\
			n_d(t+1) &= n_d(t) + \mu_n \left[V_d^t - \left( I_{\rm d}(t) - \tilde{I}_d(t)\right)^2\right] \\
			\tilde{I}_p(t+1) &= (1-\mu_{av})\tilde{I}_p(t) + \mu_{av} I_{\rm p}(t+1) \\
			\tilde{I}_d(t+1) &= (1-\mu_{av})\tilde{I}_d(t) + \mu_{av} I_{\rm d}(t+1)
		\end{align}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		Here, $I_{\rm p}^t$, $I_{\rm d}^t$, $V_p^t$ and $V_p^t$ define targets for the 
		temporal means and variances of $I_{\rm p}$ and $I_{\rm d}$. The dynamic variables 
		$\tilde{I}_p$ and $\tilde{I}_d$ are simply low-pass filtered running 
		averages of $I_{\rm p}$ and $I_{\rm d}$.

@@

		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{table}
			\begin{tabular}{ l | l || l | l }
				$\theta_{p0}$ & $0$ & $V_d^t$ & $0.25$ \\
				$\theta_{p1}$ & $-1$ & $\mu_b$ & $10^{-3}$ \\ 
				$\theta_{d}$ & $0$ & $\mu_n$ & $10^{-4}$ \\  
				$\alpha$ & $0.3$ & $\mu_{av}$ & $5 \cdot 10^{-3}$ \\   
				$\mu_w$ & $5 \cdot 10^{-5}$ & $I_{\rm p}^t$ & $0$ \\
				$V_p^t$ & $0.25$ & $I_{\rm d}^t$ & $0$  
			\end{tabular}
		\caption{Model parameters}
		\label{tab:parameters}
		\end{table}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

		\section{Results}
		\label{sect:results}

		\subsection{Increased Alignment between Basal and Apical Inputs}
		\label{sect:alignment}

		As a first test, we wanted to quantify the neuron's ability to 
		align its basal input to the apical teaching signal.
		To do so, we defined the pearson correlation coefficient
		$\rho[I_{\rm p},I_{\rm d}]$ between the basal and apical input current
		as our measure of interest. We determined this temporal
		correlation coefficient after simulating all plasticity mechanisms
		under stationary random input sequences with certain statistical
		properties that shall be explained in the following.
		As a starting point, we chose all $x_{p,i}(t)$ to be randomly
		drawn from a uniform distribution, where $x_{p,i}(t) \in (0,1)$.
		For $I_{\rm d}(t)$ to be fully `reconstructable' by the basal input, 
		$x_d(t)$ had to be some linear combination of all $x_{p,i}(t)$.
		Therefore, we chose $x_d(t) = \sum_{i=1}^N a_i x_{p,i}(t)$, where
		$\mathbf{a}$ is a random vector with unit length.
		Since we used a Hebbian learning scheme, we expected that 
		the direction and magnitude of the principal components of 
		the basal input would also significantly affect the outcome of
		the experiment: A large variance in the basal input 
		orthogonal to the `reconstruction vector' $\mathbf{a}$ 
		should act as a distraction for the plasticity and reduce the
		resulting alignment between $I_{\rm p}$ and $I_{\rm d}$. Therefore, we 
		applied a transformation to the input sequences $x_{p,i}(t)$
		that were parameterized by two quantities, a scaling factor $s$
		and the dimension $N_{dist}$ of a subspace of the basal input space.
		A set of $N_{dist}$ orthonormal basis vectors was randomly generated,
		which were also orthogonal to $\mathbf{a}$. Withing this $N_{dist}$-dimensional
		subspace, the input sequences $x_{p,i}(t)$ were then rescaled by the
		factor $s$.

		After both $x_{p,i}(t)$ and $x_d(t)$ were generated,
		a simulation was run using all previously described plasticity
		mechanisms until the dynamic variables reached a stationary state.
		After this learning phase, another set of input sequences was generated
		using the previously described protocol and $\rho[I_{\rm p},I_{\rm d}]$ was calculated.
		Note that plasticity was turned off in this phase. This entire procedure
		allowed us to calculate $\rho[I_{\rm p},I_{\rm d}]$ as a function of the distraction
		parameters $s$ and $N_{dist}$. The results for both neuron models 
		is shown in Fig.~\ref{fig:corr_dimension_scaling}. Here, the total number of 
		basal inputs was $N=10$. One can observe a decorrelation transition for 
		both models. However, the compartment model supports a significantly stronger
		distraction in terms of the scaling factor $s$ as compared to the point model.
		This was a first confirmation of our hypothesis that nonlinear interactions
		between basal and apical input could improve learning guided by top-down signals.

		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{figure}
			\includegraphics[width=\columnwidth]{corr_dimension_scaling}
			\caption{{\bf Alignment between Basal and Apical Input.} Color
			encodes the Pearson correlation $\rho[I_p,I_d]$ for different
			number of orthogonal distraction directions $N_{dist}$ 
			and the corresponding scaling faction $s$.}
			\label{fig:corr_dimension_scaling}
		\end{figure}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

		\subsection{Supervised Learning in a Linear Classification Task}

		Next, we investigated if the observed differences would also improve
		the performance in an actual supervised learning task.
		For this purpose, we constructed presynaptic basal input as illustrated
		in Fig.~\ref{fig:illustration_classification}. Written in vector form,
		each sample from the basal input was generated from the following expression:
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{equation}
			\mathbf{x}_p(t) = \mathbf{b} + \mathbf{a}\left(c(t) + \sigma_a \zeta_a(t) \right) 
			+ s \cdot \sum_{i=1}^{N_{dist}} \zeta_{dist,i}(t) \mathbf{v}_{dist,i} \; .
		\end{equation}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		Here, $\mathbf{b}$ is a random vector drawn uniformly from
		$(0,1)^N$, $\mathbf{a}$ is random unit vector as introduced in 
		Section~\ref{sect:alignment}, $c(t)$ is a binary variable drawn 
		from $\{-0.5,0.5\}$ with equal probability and $\zeta_a(t)$ and the
		$\zeta_{dist,i}(t)$ are independent random Gaussian variables.
		Finally, the set of $\mathbf{v}_{dist,i}$ forms a randomly generated
		orthogonal basis of $N_{dist}$ unit vectors which---just as in 
		Section~\ref{sect:alignment}---are also orthogonal to $\mathbf{a}$.
		As indicated by the time dependence, the Gaussian and binary random
		variables were drawn for each time step. The vectors
		$\mathbf{b}$, $\mathbf{a}$, and $\mathbf{v}_{dist,i}$ were generated
		once before the beginning of a simulation run.

		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{figure}
			\centering
			\includegraphics[width=0.75\columnwidth]{illustration_classification}
			\caption{{\bf Input Space for a Simple Linear Classification Task.}
			Two clusters of presynaptic basal activity were generated from 
			multivariate Gaussian distributions. Here, $s$ denotes the standard
			deviation orthogonal to the normal vector of the classification
			hyperplane.}
			\label{fig:illustration_classification}
		\end{figure}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
